---
title: "Final project SMDS-2-2020"
author: "Zavadskaya Katsiaryna 1847985"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(R2jags)
require(mcmcse)
require(bayesplot)
require(TeachingDemos)
require(ggmcmc)
```

## Flight Delays

The dataset contains information about flight delays for 14 airlines for each of the days of 2015. For the computations simplicity I grouped it in way that it is showing total number of flights performed and number of flights with a delay for all the airline companies during 1st of January 2015.

The original data can be accessed at:
https://www.kaggle.com/usdot/flight-delays/data?select=flights.csv


### 1 - Data

The data is shown in the table below:
```{r}
flights <- read.csv("data/flights_summary_delays_2015_01_01_01.csv")
flights <- as.data.frame(flights)
flights
```
#### 1.1 - Data overview

As we can see from the data table, the airline name is encoded with two letters. The basic summaries about our numerical features can be seen below:
```{r}
summary(flights$TOTAL)
summary(flights$DELAYED.FLIGHTS)
```
More in detail, the data available corresponds to 14 different airlines, having a minimum number of flights per day equal to 152 and a maximum of 1559. The delay rates are in the interval of minimum 39 and maximum 771 delayed flights per day.

Following we can see the graphic visualization of the data:
```{r}
# Histograms
par(mfrow=c(2,2))
hist(main='Number of flights per day', flights$TOTAL, xlab='Number of flights', ylab='Frequency', col='orchid')
hist(main='Number of delays per day', flights$DELAYED.FLIGHTS, xlab='Number of delays', ylab='Frequency', col='orchid')
d <- density(flights$DELAYED.FLIGHTS) # the density of the data
plot(d, main = "Density Delayed Flights", col = "orchid", lwd = 2)
```

### 2 - Model description

Lets denote total number of flights performed by each airline \(i\) as \(n_i \). The number of delayed flights will be \({r_i}\) which is a binary response variable with 'true' delay probability \({p_i}\). In order to model the outcome I am going to use Binomial distribution. 

<center>
\(r_i \sim Binomial(p_i,n_i) \)
</center>

The aim of my analysis is to estimate parameters of the Binomial distribution for each of the airlines.

Additionally, it is assumed that the true delay probabilities are independent for each airline. A standard non-informative prior distribution for the \(p_i\) is following:

<center>
\(p_i \sim Beta(1.0,1.0)\)
</center>

The Beta distribution is a continuous distribution that only takes values between 0 to 1.
And we need a prior for each of the \(p_i\).

In order to make inference in an automatic way, I am writing a model for JAGS.

#### 2.1 - Data preparation for JAGS

Since JAGS is not able to deal with dataframes, I need to transform data into lists.

```{r}
n <- flights$TOTAL
r <- flights$DELAYED.FLIGHTS
N <- nrow(flights)
flights.jags <- list("r", "n", "N")
```
#### 2.2 - JAGS Model

The following code was executed in order to find probabilities of delay for each airline.


```{r}
# Model
model <- function() {
  for(i in 1:N){
    p[i] ~ dbeta(1.0, 1.0) # Prior
    r[i] ~ dbinom(p[i], n[i]) # Model
  }
}

# Starting values
mod.inits = function(){
  list("p" = c(0.1, 0.1, 0.1, 0.1,0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1))
  }

# Run JAGS
set.seed(123)
mod.fit <- jags(data = flights.jags,                            
                model.file = model, inits = mod.inits,          
                parameters.to.save = 'p',                  
                n.chains = 3, n.iter = 10000, n.burnin = 1000, n.thin=5)
```

#### 2.3 - Output and Diagnostic


```{r}
mod.fit
mod.fit$BUGSoutput$summary
```

Next I want to understand if I can trust the MCMC results. To do that I want to see if there any evidence that the chains are stationary, i.e. if the simulated parameters values are moving in a consistent direction.

To get a visual idea about what is happening in a chain, I extract it from the model output.

```{r}
chainArray <- mod.fit$BUGSoutput$sims.array
```

Next I look at the traceplots in order to inspect sampling behavior and assess mixing across chains and convergence. Lets first look at all the traceplots and then have a closer view of one of them:

```{r, fig.width=16, fig.height=8}
mcmc_trace(chainArray)
mcmc_trace(chainArray, pars="p[2]")
```

While examining the following traceplots I see that chains are visualized just on top of each other, which means that all 3 chains converged towards the same value. I conclude that there can not be seen any evidence that the chain has not achieved stationarity.

The next diagnostic step is to look at the plots of estimated density of the parameters and of the deviance, where density plots are just smoothed histograms of the samples.

```{r, fig.width=16, fig.height=8}
mcmc_dens(chainArray)
mcmc_dens_overlay(chainArray)
```

The third way is to check autocorrelation plots. These plots are specific for each of the chains. I expect to see low autocorrelation, since it indicates convergence. And as it can be seen on the plot, the autocorrelation function drops to 0 pretty fast, after only a couple of lags.

```{r, fig.width=16, fig.height=8}
bayesplot::mcmc_acf(chainArray)
```

### 3 - Inferential findings

Lets first join the deviance and p[i] for all 3 chains:
```{r}
chainMat <- mod.fit$BUGSoutput$sims.matrix
```

Now the point estimates can be seen:
```{r}
p.hat.jags <- colMeans(chainMat)
p.hat.jags
```

When it comes to interval estimation, first lets see equal tail intervals:
```{r}
cred <- 0.95
p.ET.jags <- apply(chainMat, 2, quantile, prob=c((1-cred)/2, 1-(1-cred)/2))
p.ET.jags
```

And also HPD intervals:
```{r}
p.HPD.jags <- coda::HPDinterval(as.mcmc(chainMat))
p.HPD.jags
```


### 4 Alternative model

In the first model it was assumed that the true delay probabilities are independent for each airline. However, a more realistic model for flights delay is that the delay rates across airline companies are similar in some way.

<center>
\(r_i \sim Binomial(p_i,n_i) \)
</center>

<center>
\(\log(p_i/(1-p_i) \sim Normal(\mu,\tau^{2}) \)
</center>

In this model we assume that the logit of each airline's rate is related to each other.
Standard priors are given by:

<center>
\(\mu \sim Normal(0, 1*10^{-6})\)
</center>

<center>
\(1/\tau^{2} \sim Gamma(0.001, 0.001) \)
</center>

Standard non-informative priors are specified for the population mean (logit) probability of delay - \( \mu\) , and precision - \( \tau\).

#### 4.1 - Data preparation for JAGS

Since JAGS is not able to deal with dataframes, I need to transform data into lists.

```{r}
n <- flights$TOTAL
r <- flights$DELAYED.FLIGHTS
N <- nrow(flights)
flights.jags2 <- list("r", "n", "N")
```
#### 4.2 - JAGS Model

The following code was executed in order to find probabilities of delay for each airline.

```{r}
# Model
model2 <- function() {
  for(i in 1:N){
    r[i] ~ dbinom(p[i], n[i]) #Model
    b[i] ~ dnorm(mu,tau)
    logit(p[i]) <- b[i]
  }
  mu ~ dnorm(0.0, 1.0E-6) #Priors
  tau ~ dgamma(0.001, 0.001)
  
  sigma <- 1 / sqrt(tau)
  pop.mean <- exp(mu) / (1 + exp(mu))
}

# Starting values
mod.inits2 = function(){
  list(b = c(0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1, 0.1),
       tau = 1,
       mu = 0)
}

# Run JAGS
set.seed(123)
mod.fit2 <- jags(data = flights.jags2,                            
                model.file = model2, inits = mod.inits2,          
                parameters.to.save = c("p","sigma","mu","pop.mean"),                  
                n.chains = 3, n.iter = 10000, n.burnin = 1000, n.thin=5)
```
#### 4.3 - Output and Diagnostic


```{r}
mod.fit2
mod.fit2$BUGSoutput$summary
```

Next I want to understand if I can trust the MCMC results. To do that I repeat the process exactly as for model 1.

```{r}
chainArray2 <- mod.fit2$BUGSoutput$sims.array
```

Traceplots:

```{r, fig.width=16, fig.height=8}
mcmc_trace(chainArray2)
mcmc_trace(chainArray2, pars="p[3]")
```

While examining the following traceplots I see that chains are visualized just on top of each other, which means that all 3 chains converged towards the same value. I conclude that there can not be seen any evidence that the chain has not achieved stationarity.

Density:

```{r, fig.width=16, fig.height=8}
mcmc_dens(chainArray2)
mcmc_dens_overlay(chainArray2)
```

Autocorrelation plots:

```{r, fig.width=16, fig.height=8}
bayesplot::mcmc_acf(chainArray)
```

### 5 - Inferential findings of the model 2

```{r}
chainMat2 <- mod.fit2$BUGSoutput$sims.matrix
```

Now the point estimates can be seen:
```{r}
p.hat.jags2 <- colMeans(chainMat2)
p.hat.jags2
```

Equal tail intervals:
```{r}
cred <- 0.95
p.ET.jags2 <- apply(chainMat2, 2, quantile, prob=c((1-cred)/2, 1-(1-cred)/2))
p.ET.jags2
```

HPD intervals:
```{r}
p.HPD.jags2 <- coda::HPDinterval(as.mcmc(chainMat2))
p.HPD.jags2
```
### 5 - Comparison of the models

In order to compare two different models I need to use term DIC - Deviance Information Criterion. DIC is a Bayesian method for model comparison and a measure for the prediction error of a model. The lower DIC value the better the model fits the data. 

The deviance D, given a parameter value $\theta$ returns error using the log-likelihood function:
$D(\theta) = -2  \text{log} * L(data|\theta)$

Once the posterior means are computed, I can measure their deviance as:

$D(\hat{\theta}) = -2 \text{log} * L(data|\hat{\theta})$
Taking all steps i in the different MCMCs, we can from this compute an average - posterior mean of the deviance as follows:

$\hat{D} = \frac{1}{M}\sum^{M}_{i=1}D (\theta_{i})$

Now it is possible to compute the effective number of parameters:

$p_{D} = \hat{D} - D(\hat{\theta})$

The DIC is then computed:

$DIC = D(\hat{\theta})- 2 p_{D}$

```{r}
paste("Model 1 DIC:",round(mod.fit$BUGSoutput$DIC,digits = 3),'         ',"Model 2 DIC",round(mod.fit2$BUGSoutput$DIC,digits = 3))
```
Looking at the DIC values, it is obvious that the difference is not significant. And in fact the difference is equal to:

```{r}
paste("DIC Difference:", round(abs(mod.fit$BUGSoutput$DIC - mod.fit2$BUGSoutput$DIC),digits = 3))
```
The conclusion I can draw is that it is not possible to say exactly which model out of two is better.

### 6 - Comparison with frequentist inference

Instead of using MCMC approximation for estimation of each of the probabilities, it is also possible to estimate CI of their mean and standard deviation using Bootstarp.

Firstly, I compute probabilities by dividing column 'DELAYED.FLIGHTS' by column 'TOTAL':
```{r}
flights$PROB =  flights$DELAYED.FLIGHTS / flights$TOTAL
flights
```

At first number of sampling was picked. After that in the loop list of estimated probabilities that we calculated below was resampled with replacement.
Mean and Standard deviation of each sample were saved. Using percentiles we can construct our Confidence Intervals.

```{r}
PROB_Matrix <- as.matrix(flights$PROB)
n = nrow(flights)
B <- 1000

our_out_put <- rep(NA,B) # empty vector
our_out_sd <- rep(NA,B)

for (b in 1:B){
  new_sample <- PROB_Matrix[sample(n,n,replace=TRUE),]
  our_out_put[b] <- mean(new_sample)
  our_out_sd[b] <- sqrt(var(new_sample))
}

paste("Mean Lower Bound:", quantile(our_out_put, c(0.025)),"Mean Upper Bound:",quantile(our_out_put, c(0.975)))
paste("SD Lower Bound:", quantile(our_out_sd, c(0.025)),"SD Upper Bound:",quantile(our_out_sd, c(0.975)))
```





